---
title: "3. Fitting models and making predictions"
output: rmarkdown::html_vignette
bibliography: biblio.bib
vignette: >
  %\VignetteIndexEntry{3. Fitting models and making predictions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spmismo)
library(ggplot2)
library(sf)
```

If interested in the theory involving this calculations, take a look at this
[vignette](https://lcgodoy.me/spmismo/articles/theory.html).

# Application

To exemplify the fitting and predicting functions, we are going to use the data
from @johnson2020dealing again^[for more details about the data, take a look at
the vignette [sf-to-spm](https://lcgodoy.me/spmismo/articles/sf-to-spm.html)]. 
First, in the code chunk below, we load the data and convert it to `spmismo` 
compatible classes. Since this is a toy example, we are going to generate a 500 
points grid for the MSOA data and 5 points by polygon for the LSOA data.


```{R load-data}
data(liv_msoa)
data(liv_lsoa)

## msoa to spm
spm_msoa <-
    sf_to_spm(sf_obj1 = liv_msoa,
              n_pts   = 500,
              type    = "regular",
              by_polygon = FALSE,
              poly_ids = "msoa11cd",
              var_ids  = "MLE")

## lsoa to spm
spm_lsoa <-
    sf_to_spm(sf_obj1 = sf_b,
              n_pts   = 5,
              type    = "regular",
              by_polygon = TRUE,
              poly_ids = "lsoa11cd",
              var_ids  = "IMDScore")
```

## Fit at MSOA, prediction at LSOA


```{R fit-pred-1}
theta_st_lsoa <-
    c("alpha" = 70, "omega" = 1,
      "sigsq" = 1, "phi" = 1)

fit_msoa <-
    fit_spm(x = spm_lsoa,
            theta_st = theta_st_lsoa,
            model = "pexp",
            kappa = 1.5,
            apply_exp = TRUE,
            control   = list(maxit = 500))

summary_sspm_fit(fit_y)

pred_y <- predict_spm(x = fit_y)

plot(pred_y$pred_grid[c("mu_pred", "se_pred")],
     pch = 19)

pred_y <- predict_spm(x = sf_b, spm_obj = fit_y,
                      .aggregate = TRUE, n_pts = 5000,
                      type = "hexagonal")

plot(pred_y$pred_agg, pch = 15)

##--- testing x ----



plot(spm_x$grid, pch = 15)

fit_x <-
    fit_spm(x = spm_x,
            theta_st  = c("alpha" = 25, "omega" = 1,
                          "sigsq" = 1, "phi" = 1),
            model = "matern",
            kappa = .5,
            apply_exp = TRUE,
            control   = list(maxit = 500))

summary_sspm_fit(fit_x)

pred_x <- predict_spm(x = fit_x)

plot(pred_x$pred_grid[c("mu_pred", "se_pred")],
     pch = 19)

pred_x <- predict_spm(x = sf_a, spm_obj = fit_x,
                      .aggregate = TRUE, n_pts = 5000,
                      type = "regular")

plot(pred_x$pred_grid[c("mu_pred", "se_pred")], pch = 15)

plot(pred_x$pred_agg)

##--- testing  x, y ----

plot(spm_xy$x$grid, pch = 15)

plot(spm_xy$y$grid, pch = 15)

theta_xy <-
    c("alpha" = 60, "beta" = -1,
      "gamma" = 30, "omega" = 50,
      ## "omega2" = .5, "omega3" = 1,
      "sigsq" = 5, "phi" = 2,
      "nusq" = 100)

debugonce(spmismo:::mult_log_lik)

aa <-
    spmismo:::mult_log_lik(theta = theta_xy,
                           .dt = list(spm_xy$y$var,
                                      spm_xy$x$var),
                           dists = list(spm_xy$y$dists,
                                        spm_xy$x$dists,
                                        spm_xy$cdist),
                           model = "pexp",
                           kappa = 1.5,
                           apply_exp = TRUE)


p <- 1
(2 * p + 4 + (0.5 * p * (p + 1)))
length(theta_xy)

fit_xy <-
    fit_spm(x = spm_xy,
            theta_st  = theta_xy,
            model = "matern",
            kappa = .5,
            apply_exp = TRUE,
            control   = list(maxit = 700))
```


## Y | X

```{R spm-msoa-lsoa}
## both to spm
spm_xy <-
    sf_to_spm(sf_obj1 = sf_a,
              sf_obj2 = sf_b,
              n_pts   = list(1700, 5),
              type    = "regular",
              by_polygon = c(FALSE, TRUE),
              poly_ids = list("msoa11cd", "lsoa11cd"),
              var_ids  = list("MLE", "IMDScore"))
```

# Reference
