---
title: "3. Fitting models and making predictions"
output: rmarkdown::html_vignette
bibliography: biblio.bib
vignette: >
  %\VignetteIndexEntry{3. Fitting models and making predictions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(smile)
library(ggplot2)
library(sf)
```

If interested in the theory involving this calculations, take a look at this
[vignette](https://lcgodoy.me/smile/articles/theory.html).

# Application

To exemplify the fitting and predicting functions, we are going to use the data
from @johnson2020dealing again^[for more details about the data, take a look at
the vignette [sf-to-spm](https://lcgodoy.me/smile/articles/sf-to-spm.html)]. 
First, in the code chunk below, we load the data and convert it to `smile` 
compatible classes. Since this is a toy example, we are going to generate a 500 
points grid for the MSOA data and 5 points by polygon for the LSOA data.

```{R load-data}
data(liv_msoa)
data(liv_lsoa)

## workaround for compatibility with different PROJ versions
st_crs(liv_msoa) <-
    st_crs(liv_msoa)$input

st_crs(liv_lsoa) <-
    st_crs(liv_lsoa)$input

## msoa to spm
spm_msoa <-
    sf_to_spm(sf_obj = liv_msoa,
              n_pts  = 500,
              type   = "regular",
              by_polygon = FALSE,
              poly_ids = "msoa11cd",
              var_ids  = "leb_est")

## lsoa to spm
spm_lsoa <-
    sf_to_spm(sf_obj = liv_lsoa,
              n_pts  = 5,
              type   = "regular",
              by_polygon = TRUE,
              poly_ids = "lsoa11cd",
              var_ids  = "imdscore")
```

The map below shows the LEB observed at the MSOAs.
```{R leb-msoa1}
ggplot(data = liv_msoa,
       aes(fill = leb_est)) +
    geom_sf(color = 1, lwd = .1) +
    scale_fill_viridis_c(option = "H") +
    theme_minimal()
```

The map below displays the IMD observed at the LSOAs.
```{R imd-lsoa1}
ggplot(data = liv_lsoa,
       aes(fill = imdscore)) +
    geom_sf(color = 1, lwd = .1) +
    scale_fill_viridis_c(option = "H") +
    theme_minimal()
```

## Fit at MSOA, prediction at LSOA

Using as starting parameters $\alpha = 70$ (average of $Y$), $\omega = 1$ (can
be seen as a nugget effect), $\sigma^2$ (variance parameter for the powered
exponential family), and $\phi = 1$ (paramter that controls the range os spatial
dependence, 1 here represents 1_km_). We are going to compute the Maximum
Likelihood estimates considering a Powered Exponential covariance function to
the data, with $\kappa = 1.5$^[This parameter is known to be hard to be
estimated in the literature of geostatistics.].  Note that, it is important for
this inputing vector to be a named vector, because some functions track the
parameters by their names. 
```{R fit-pred-1}
theta_st_msoa <- c("phi" = 1)

fit_msoa <-
    fit_spm(x = spm_msoa,
            theta_st = theta_st_msoa,
            model = "pexp",
            kappa = .5,
            apply_exp = TRUE,
            opt_method = "L-BFGS-B",
            control   = list(maxit = 500))
```

After a few seconds, we get the Results. Now, we can use the function
`summary_sspm_fit` to get the summary of the results. The function prints a
table containing the estimates, their standard errors, and `1 - sig` confidence
interval based on the asymptotic normality of the MLEs. The function also prints
a message indicating if the optimization algorithm has converged^[The default
algorithm used by the package is the Nelder-Mead, this can easily be changed by
setting the parameter `opt_method = "L-BFGS-B"`, for example.].
```{R summary-fit1}
summary_spm_fit(fit_msoa, sig = .05)
```

Finally, now we are going to perform predictions of the LEB at the LSOA
level. To do so, we use the function `predict_spm`. The `x` argument gives the
sf on which we want to make the predictions according to the `sbm_obj`, which
represents the estimated parameters. Now, take a look at the remaining
arguments. The last two, are analogous to what we have seen for the function
`sf_to_spm`. They are concerning the grid generated over the study region, now
we going to make predicitons over this grid. I set it to 2500 to "approximate" a
continuous region^[Be careful here, a very large number of points can result in
intractable covariance matrices]. The argument `.aggregate` if set to `TRUE`,
computes the predictions over the polygons from the input `x` (in this case
`liv_lsoa`). This function also works if the input is a `sf` object with POINT
geometries. The objects returned by the function `pred_lsoa` are lists with the
following elements: 1) `mu_pred` - a vector with the predicted average for each
location in the generated grid; 2) `sig_pred` the variance covariance matrix
associated with the same grid; 3) the predictions (and standard errors)
organized in an `sf` object to facilitate the visualization of the predictions
over a (almost) continuous space; and 4) `pred_agg` which is returned only if we
set `.aggregate = TRUE`. The last one contains the predictions and their
standard erros integrated over the "new regions".
```{R predict-fit1}
pred_lsoa <- predict_spm(x = liv_lsoa, spm_obj = fit_msoa,
                         .aggregate = TRUE, n_pts = 2500,
                         type = "regular")
```

Now, let's make some data manipulation to make the visualization look nicer. We
are going to stack all the data to make it easier to plot them all together
using ggplot2. The next map whos the predictions.
```{R manip-pred-1}
aux_viz1 <- vector(mode   = "list",
                   length = 4)

aux_viz1[[1]] <- transform(pred_lsoa$pred_grid["mu_pred"],
                           surface  = "continuous",
                           variable = "prediction")
names(aux_viz1[[1]])[1] <- "value"
aux_viz1[[2]] <- transform(pred_lsoa$pred_grid["se_pred"],
                           surface  = "continuous",
                           variable = "se")
names(aux_viz1[[2]])[1] <- "value"
aux_viz1[[3]] <- transform(pred_lsoa$pred_agg["mu_pred"],
                           surface  = "aggregated",
                           variable = "prediction")
names(aux_viz1[[3]])[1] <- "value"
aux_viz1[[4]] <- transform(pred_lsoa$pred_agg["se_pred"],
                           surface  = "aggregated",
                           variable = "se")
names(aux_viz1[[4]])[1] <- "value"

aux_viz1 <- do.call("rbind", aux_viz1)

ggplot(data = aux_viz1[aux_viz1$variable == "prediction",],
       aes(fill  = value,
           color = value)) +
    geom_sf(data = st_geometry(pred_lsoa$pred_agg), color = 1,
            lwd = .1, inherit.aes = FALSE) +
    geom_sf(pch = 15) +
    scale_fill_viridis_c(option = "H") +
    scale_color_viridis_c(option = "H") +
    guides(color = FALSE) +
    facet_wrap(surface ~ .) +
    theme_bw()
```

The next one displays the standard errors for the predictions. Clearly, we have
less variability in the neighborhoods on which we have small polygons.
```{R se-ggplot1}
ggplot(data = aux_viz1[aux_viz1$variable == "se",],
       aes(fill  = value,
           color = value)) +
    geom_sf(data = st_geometry(pred_lsoa$pred_agg), color = 1,
            lwd = .1, inherit.aes = FALSE) +
    geom_sf(pch = 15) +
    scale_fill_viridis_c(option = "H") +
    scale_color_viridis_c(option = "H") +
    guides(color = FALSE) +
    facet_wrap(surface ~ .) +
    theme_bw()
```

```{R rm-obj1, echo = TRUE}
rm(aux_viz1, pred_lsoa, fit_msoa)
invisible(gc(full = TRUE))
```

## Fit at LSOA, predict at MSOA

Using as starting parameters $\alpha = 25$ (average of $Y$), $\omega = 1$ (can
be seen as a nugget effect), $\sigma^2$ (variance parameter for the powered
exponential family), and $\phi = 1$ (paramter that controls the range os spatial
dependence, 1 here represents 1_km_). We are going to compute the Maximum
Likelihood estimates considering a Matern covariance function to the data, with
$\kappa = 2.5$.

```{R fit-pred-2}
theta_st_lsoa <-
    c("phi" = 1)

fit_lsoa <-
    fit_spm(x = spm_lsoa,
            theta_st = theta_st_lsoa,
            model = "matern",
            kappa = .5,
            apply_exp = TRUE,
            opt_method = "L-BFGS-B",
            control   = list(maxit = 500))
```

After a few seconds, we get the Results. Now, we can use the function
`summary_sspm_fit` again to summarise the estimated parameters. 
```{R summary-fit2}
summary_spm_fit(fit_lsoa, sig = .05)
```

Now we are going to perform predictions of the IMD at the MSOA level. Since from
now on the process is pretty much the exam, the explanation will be shorter.
```{R predict-fit2}
pred_msoa <- predict_spm(x = liv_msoa, spm_obj = fit_lsoa,
                         .aggregate = TRUE, n_pts = 2500,
                         type = "regular")
```

Now, let's make some data manipulation to again. Note that, the predictions for
this variable have more
```{R manip-pred-2}
aux_viz2 <- vector(mode   = "list",
                   length = 4)

aux_viz2[[1]] <- transform(pred_msoa$pred_grid["mu_pred"],
                           surface  = "continuous",
                           variable = "prediction")
names(aux_viz2[[1]])[1] <- "value"
aux_viz2[[2]] <- transform(pred_msoa$pred_grid["se_pred"],
                           surface  = "continuous",
                           variable = "se")
names(aux_viz2[[2]])[1] <- "value"
aux_viz2[[3]] <- transform(pred_msoa$pred_agg["mu_pred"],
                           surface  = "aggregated",
                           variable = "prediction")
names(aux_viz2[[3]])[1] <- "value"
aux_viz2[[4]] <- transform(pred_msoa$pred_agg["se_pred"],
                           surface  = "aggregated",
                           variable = "se")
names(aux_viz2[[4]])[1] <- "value"

aux_viz2 <- do.call("rbind", aux_viz2)

ggplot(data = aux_viz2[aux_viz2$variable == "prediction",],
       aes(fill  = value,
           color = value)) +
    geom_sf(data = st_geometry(pred_msoa$pred_agg), color = 1,
            lwd = .1, inherit.aes = FALSE) +
    geom_sf(pch = 15) +
    scale_fill_viridis_c(option = "H") +
    scale_color_viridis_c(option = "H") +
    guides(color = FALSE) +
    facet_wrap(surface ~ .) +
    theme_bw()
```

The next one displays the standard errors for the predictions. 
```{R se-ggplot2}
ggplot(data = aux_viz2[aux_viz2$variable == "se",],
       aes(fill  = value,
           color = value)) +
    geom_sf(data = st_geometry(pred_msoa$pred_agg), color = 1,
            lwd = .1, inherit.aes = FALSE) +
    geom_sf(pch = 15) +
    scale_fill_viridis_c(option = "H") +
    scale_color_viridis_c(option = "H") +
    guides(color = FALSE) +
    facet_wrap(surface ~ .) +
    theme_bw()
```

# Reference
