---
title: "3. Areal Interpolation"
output: rmarkdown::html_vignette
bibliography: biblio.bib
vignette: >
  %\VignetteIndexEntry{3. Areal Interpolation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Often, in an applied setting, it is interesting to compute estimates (and
quantify its associated uncertainty) of socioeconomic and demographic variables
for different administrative regions within the same city. Computing estimates
is trivial when applying the Areal Interpolation [@goodchild1980areal]
approach. However, a common criticism regarding this method is that it does not
provide uncertainty measures associated with the estimates
[@bradley2016bayesian]. Our package provides the simple areal interpolation
along with an approximation and an upper bound for the variance of the computed
estimates when the variance associated with the observed data is known, which is
a common fact when working with survey data. To illustrate how does it work, we
are going to show how to estimate a variable observed at census tracts in the
Public Use Microdata Areas (PUMAs) in NYC.

To load the data and the packages needed for this tutorial run the chunk of code
below.
```{r load-data, eval = TRUE}
library(sf)
library(smile)
library(ggplot2)

data(nyc_surv)
data(nyc_comd)
```

Since the `nyc_surv` data proves a "margin of error" instead of a variance, we
are going to use the piece of code below to approximate the variance. We are
assuming the margin of error was calculated using a 95% confidence interval
based on the Normal distribution.
```{r var_calc}
nyc_surv <- transform(nyc_surv,
                      my_var = moe / qnorm(p = .975))
nyc_surv <- transform(nyc_surv, my_var = my_var * my_var)
```
Now, to estimate a variable observed at the `nyc_surv` dataset into the
`nyc_comd` data we can run the following command:
```{r est-1, eval = FALSE}
estimate_comd <-
    ai(source = nyc_surv, target = nyc_comd,
       vars = "estimate")
```
where `source` is the "source" dataset, i.e., the one on which we observe the
variable we wish to estimate at the `target` dataset. The argument `vars` is a
`character` (scalar or vector) containing the name of the variable (or
variables) at the `source` that are going to be estimated at the `target`
regions. This function does not quantify the uncertainty associated with the
estimates, though. In order to get a variability measure associated with the
estimates, we need to use the function `ai_var` (see the chunk of code
below). The function is quite similar to the before mentioned `ai`. The
differences are: 1) `vars` must be a scalar; 2) the argument `my_var` is another
length 1 `character` indicating the name of the variable that contains the
variance of the observed variable in the `source` dataset; and 3) the argument
`var_method` indicates whether we are going to use the Moran's I (`var_method =
"MI"`) to approximate the autocorrelation between areas and, therefore,
approximate the variance of the estimates or the Cauchy--Schwartz (`var_method =
"CS"`) inequality to calculate an upper bound for such variances. The output of
the `ai_var` function is the `target` (`sf data.frame`) with two additional
columns, `est` and `se_est`, representing the estimated variable and its
corresponding (approximate) standard error.
```{r est-2}
estimate_comd <-
    ai_var(source = nyc_surv, target = nyc_comd,
           vars = "estimate", vars_var = "my_var",
           var_method = "MI")
```
The next piece of code uses `ggplot2` to plot the observed and estimated
variable at NYC census tracts and community districts, respectively.
```{r ggplot-obs-est}
viz_dt <-
    rbind(
        transform(nyc_surv, source = "Observed", est = estimate)[c("source", "est")],
        transform(estimate_comd, source = "Predicted")[c("source", "est")]
    )

ggplot(data = viz_dt,
       aes(fill = est)) +
    geom_sf(color = 1, lwd = .1) +
    scale_fill_viridis_c(option = "H") +
    theme_bw() +
    facet_wrap( ~ source) +
    theme(axis.text = element_blank())
```

```{r ggplot-obs-est-viz, eval = TRUE, echo = FALSE}
data(viz_dt)

ggplot(data = viz_dt,
       aes(fill = estimate / 1000)) +
    geom_sf(color = 1, lwd = .1) +
    scale_fill_viridis_c(option = "H",
                         name = "median \n income",
                         label = \(x) paste0(x, "K")) +
    theme_bw() +
    facet_wrap( ~ source) +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank())
```


# Reference
